Data-driven Reduction of Multiscale Stochastic Dynamical Systems \thanks{This work was supported by ....}

### 2015-05-28 reducible_SDE/paper/SIADS_paper/SIADS_revisions.tex
added some edits to SIADS revisions

Following the methods of Mori \cite{mori1965transport}, Zwanzig \cite{zwanzig1961memory}, and others \cite{brey1981nonlinear, chorin2000optimal, hijon2010mori}, one can reduce the number of variables needed to describe a system of differential equations.
%
However, in general, this reduction introduces memory terms.
%
It transforms a system of differential equations into a system of (lower-dimensional) {\em integro-}differential equations,
so that the reduction of the number of variables is counterpoised by the increased complexity of the  reduced model.
%
Here, we will study the special case of evolution equations which contain an inherent time scale separation;
in this case, it is possible, in principle, to obtain a reduced system of differential equations in only the slow variables {\em without memory terms}.
%
Such an analysis crucially hinges on knowing in which variables (or, more generally, functions of variables) one can write such a reduced system of slow evolution equations.

@@

Given an explicit system of ordinary differential equations,
one can make numerical approximations, such as the quasi-steady state approximation \cite{segel1989quasi} or
the partial equilibrium approximation \cite{gallagher1986combined}, to reduce the system dimensionality without introducing memory terms.
%
There has been some recent analytical work on extending and
generalizing such ideas to more complex systems of equations \cite{ait2008closed, calderon2007fitting, contou2011model, dong2007simplification, givon2004extracting, pavliotis2007parameter,  sotiropoulos2009model}.
%
However, in many instances, closed form, analytical models are not given explicitly,
but can only be inferred from simulation and/or experimental data.
%
We therefore turn to data-driven techniques to analyze such systems and uncover the relevant dynamical modes.
%
In particular, we will use a manifold-learning based approach, as such methods can accommodate nonlinear structures in high-dimensional data.

@@

We will present detailed analysis for our method, and provide conditions under which it will successfully recover the slow variables.
%
Furthermore, based on this analysis, we will present data-driven protocols to tune the parameters of the method appropriately.
%
Our presentation and discussion will address two-time-scale stochastic systems; however, we claim that
our framework and analysis readily extends to systems with multiple time scale separations.
%

@@

Consider the following two-time-scale system of stochastic differential equations (SDEs),
\begin{equation} \label{eq:general_SDE}
\begin{aligned}
dx_i(t) &= a_i(\mathbf{x}(t)) dt + dW_i(t), & \: 1 \le i \le m \\
dx_i(t) &= \frac{a_i(\mathbf{x}(t))}{\epsilon} dt + \frac{1}{\sqrt{\epsilon}} dW_i(t) , & \: m+1 \le i \le n
\end{aligned}
\end{equation}
where $W_i(t)$ are independent standard Brownian motions, $\mathbf{x}(t)  = \begin{bmatrix} x_1(t) & \cdots & x_n(t) \end{bmatrix}^T \in \mathbb{R}^n$, and $\epsilon \ll 1$.
%
In the simple case of a linear drift function, i.e., when $a_i(\mathbf{x}(t)) = \mu _i x_i$ with $\mu_i < 0$, the probability density function of $x_i$ approaches a Gaussian with
(finite) variance $\mu_i$.
%
The time constant of the approach of the variance to equilibrium is $-1/\mu_i$ for $i=1,\ldots,m$ and $-\epsilon/\mu_i$ for $i=m+1,\ldots,n$ \cite{lelievre2013optimal}.
%
Thus, the last $n-m$ variables of \eqref{eq:general_SDE} rapidly approach a local equilibrium measure and exhibit fast dynamics, while the first $m$ variables exhibit slow dynamics.
%
A short burst of simulation will yield a cloud of points which is broadly distributed in the fast directions but narrowly distributed in the slow ones.
%
With the appropriate conditions on $a_i(\mathbf{x})$, the same can be said for more general drift
functions, where $\mu_i$ are the eigenvalues of the Jacobian of $\mathbf{a}(\mathbf{x}) = \begin{bmatrix} a_1(\mathbf{x}) & \cdots & a_n(\mathbf{x}) \end{bmatrix}^T$ \cite{villani2009hypocoercivity}.
%
Therefore, \eqref{eq:general_SDE} defines an $n$-dimensional stochastic system with $m$ slow
variables and $n-m$ fast variables, and $\epsilon$ defines the time scale separation.
%
The ratio of the powers of $\epsilon$ in the drift and diffusion terms in \eqref{eq:general_SDE} is essential,
as we require the square of the diffusivity to be of the same order as the drift as $\epsilon \rightarrow 0$ \cite{berglund2003geometric}.
%
If the diffusivity is larger, then, as $\epsilon \rightarrow 0$, the equilibrium measure will be
unbounded.
%
Conversely, if the diffusivity is smaller, the equilibrium measure will go to $0$ as $\epsilon \rightarrow 0$.

Assuming the sample average of $a_i(\mathbf{x})$ converges to a distribution which is only a function of the slow variables, then by the averaging principle \cite{freidlin2012random}, we can write a reduced SDE in {\em only} the slow variables $x_1, \dots, x_m$.
%
The aim of our work is to show how we can detect such slow variables {\em automatically} from data, in order to help inform modeling efforts and aid in the writing of such reduced stochastic models.
%
In general, we are not given the variables $\mathbf{x}(t)$ from the original SDE system, but instead, we are given some {\em observations}
 in the form $\mathbf{y}(t) = \mathbf{f} (\mathbf{x}(t))$.
%
We assume that $\mathbf{f}: \mathbb{R}^n \mapsto \mathbb{R}^d$, $n \le d$, is a deterministic (possibly nonlinear) function whose image is an $n$-dimensional manifold $\mathcal{M}$ in $\mathbb{R}^d$.
%
% TODO: check whether the inverse condition is not too strong (and we only need it to apply to the domain of the samples)
%
For our analysis, we require $\mathbf{g} = \mathbf{f} ^{-1}$ to be well-defined on $\mathcal{M}$, and both $\mathbf{f}$ and $\mathbf{g}$ to be continuously differentiable to fourth order.
%
Given data $\mathbf{y}(t_1),\ldots,\mathbf{y}(t_N)$ on $\mathcal{M}$ we would like to recover a parametrization of the data that is one-to-one with the
slow variables $x_1, \dots, x_m$.

@@

If two points $\mathbf{x}(t_1)$ and $\mathbf{x}(t_2)$ are drawn from an $n$-dimensional
Gaussian distribution with covariance $\mathbf{C}_x$, the Mahalanobis distance between the points is defined as \cite{mahalanobis1936generalized}
\begin{equation}
	\| \mathbf{x}(t_1) - \mathbf{x}(t_2) \| _M = \sqrt{ (\mathbf{x}(t_1) - \mathbf{x}(t_2))^T \mathbf{C}_x^{-1} (\mathbf{x}(t_1) - \mathbf{x}(t_2) )  }.
\end{equation}
In particular, 
%%%YGK+CDS if $\mathbf{x}(t_1)$ and $\mathbf{x}(t_2)$ are samples 
for \eqref{eq:general_SDE}, whose covariance does not depend on $\mathbf{x}$,  $\mathbf{C}_x^{-1} = \mathrm{diag}(e_1, \ldots, e_n)$ is a constant matrix where
\begin{equation} \label{eq:e_def}
\begin{aligned}
e_i =& 1, \: & 1 \le i \le m \\
e_i =& \epsilon, \: & m+1 \le i \le n,
\end{aligned}
\end{equation}
and the Mahalanobis distance between samples is
\begin{equation} \label{eq:rescale_x_dist}
\| \mathbf{x}(t_2) - \mathbf{x}(t_1) \|^2_M = \sum_{i=1}^n e_i \left( x_i(t_2) - x_i(t_1) \right)^2.
\end{equation}
Note that in \eqref{eq:rescale_x_dist}, the fast variables are collapsed and become $\mathcal{O}(\sqrt{\epsilon})$ small,
and so this metric is implicitly insensitive to variations in the fast variables.
%
The metric \eqref{eq:rescale_x_dist} can be rewritten as
\begin{equation} \label{eq:norm_z}
\| \mathbf{x}(t_2) - \mathbf{x}(t_1) \|^2_M = \| \mathbf{z}(t_2) - \mathbf{z}(t_1) \|^2_2
\end{equation}
where
\begin{equation} \label{eq:general_rescale}
z_i(t) = \sqrt{e_i} x_i(t).
\end{equation}
$\mathbf{z}(t)$ is a stochastic process of the same dimension as $\mathbf{x}(t)$, rescaled so that each variable has unit diffusivity.
%
This rescaling transforms our problem from one of detecting the slow variables within dynamic data to one of traditional data mining.
%
The Mahalanobis distance incorporates information about the dynamics and relevant time scales, so that using traditional data mining techniques with this metric will allow us to detect the slow variables in our data \cite{singer2009detecting}.
%
It is important to note that, in practice, we {\em never construct} $\mathbf{z}(t)$ {\em explicitly}.
%
It was shown in \cite{singer2008non} that, assuming $\mathbf{f}$ is bilipschitz, the Mahalanobis distance can be extended to approximate (to fourth order) the Euclidean distance between the rescaled samples $\mathbf{z}(t)$ from accessible $\mathbf{y}(t) = \mathbf{f} (\mathbf{x}(t))$,
%
\begin{equation} \label{eq:mahalanobis}
\| \mathbf{y}(t_2) - \mathbf{y}(t_1) \|^2_M = \| \mathbf{z}(t_2) - \mathbf{z}(t_1) \|^2_2 + \mathcal{O}(\| \mathbf{y}(t_2) - \mathbf{y}(t_1) \|^4_2).
\end{equation}
%
This approximation is accurate when $\| \mathbf{y}(t_2) - \mathbf{y}(t_1) \|$ is small.
%
Because we will integrate these distances into a manifold learning algorithm which only considers local distances, we can recover a parametrization of the data which is consistent with the underlying system variables $\mathbf{x}(t)$, even when the data are obscured by a function $\mathbf{f}$.
%
In Section~\ref{sec:analysis}, we will show how we can approximate this distance directly from data $\mathbf{y}(t)$.
%

@@

Following \cite{singer2008non}, we will show via Taylor expansion that the Mahalanobis distance between the observations \eqref{eq:mahalanobis_distance} approximates the Euclidean
distance {\em in the rescaled variables} for general nonlinear observation functions $\mathbf{f}$ (provided $\mathbf{f}$ is bilipschitz and both $\mathbf{f}$ and $\mathbf{f}^{-1}$ are differentiable to fourth order).
%Our approximation of the pairwise distances relies on the Taylor expansion of the measurement function $\mathbf{f}$.
%
\eqref{eq:mahalanobis_distance} cannot be evaluated directly since we do not have access to the covariance matrices, so we will instead estimate the covariances directly from data.
%
We can estimate the covariance $\mathbf{C}(\mathbf{y}(t_0))$ empirically from a set of values $\mathbf{y}(t_1), \dots, \mathbf{y}(t_q)$ drawn from the local distribution at $\mathbf{y}(t_0)$.
%
One way to obtain such a set of points is to run $q$ simulations for a short time, $\delta t$, each starting from $\mathbf{y}(t_0)$.
%
Alternatively, we can consider a single time series of length $q \delta t$ starting from $\mathbf{y}(t_0)$, and then estimate the covariance
from the increments $\Delta \mathbf{y}(t_i) = \mathbf{y}(t_i) -\mathbf{y}(t_{i-1})$.
%
Although we will present analysis and results for the first type of estimation, the second case is often more practical in practice.
%
%It can be shown that if $q \delta t$ is sufficiently small, the errors introduced are of sufficiently low order.

Errors in our estimation of the Mahalanobis distance arise from three sources.
%
One source of error is approximating the function $\mathbf{f}$ locally as a linear function by truncating the Taylor expansion of $\mathbf{f}$ at first order.
%
An additional source of error arises from disregarding the drift in the stochastic process, and assuming that samples are drawn from a Gaussian distribution.
%
The third source comes from finite sampling effects.
%
In this work, we will address and discuss the first two sources of error (the finite sampling effects are the subject of future research).
%
We can control the effects of the errors due to truncation of the Taylor expansion by adjusting $\sigma_{kernel}$; the higher-order terms in this expansion will be small for points which are close, such that adjusting $\sigma_{kernel}$ will allow us to only consider distances which are
sufficiently accurate in our overall computation scheme.
%
Furthermore, we can control the errors incurred by disregarding the drift by adjusting the time scale of our simulation bursts $\delta t$.
%
Figure~\ref{fig:schematic} illustrates some of the issues in choosing the sizes $\delta t$ (or $q \delta t$ if the alternate method is used) and the parameter $\sigma_{kernel}$.
%
We will present both analytical results for the error bounds, as well as an empirical methodology to set the parameters $\sigma_{kernel}$ and $\delta t$ for our method to accurately recover the slow variable(s).

\subsection{Error due to the observation function $\mathbf{f}$}

>>>>

Following the methods of Mori \cite{mori1965transport}, Zwanzig \cite{zwanzig1961memory}, and others \cite{brey1981nonlinear, chorin2000optimal, hijon2010mori}, one can reduce the number of variables needed to describe a system of differential equations.
%
However, in general, this reduction introduces memory terms \cite{roberts2008normal}.
%
It transforms a system of differential equations into a system of (lower-dimensional) {\em integro-}differential equations,
so that the reduction of the number of variables is counterpoised by the increased complexity of the  reduced model.
%
Here, we will study the special case of evolution equations which contain an inherent time scale separation;
in this case, it is possible (for weak models or for models with linear noise) to obtain a reduced system of differential equations in only the slow variables {\em without memory terms} \cite{roberts2008normal}.
%
Such an analysis crucially hinges on knowing in which variables (or, more generally, functions of variables) one can write such a reduced system of slow evolution equations.

@@

Given a system of ordinary differential equations with an explicit functional form,
one can make numerical approximations, such as the quasi-steady state approximation \cite{segel1989quasi} or
the partial equilibrium approximation \cite{gallagher1986combined}, to reduce the system dimensionality.
%
There has been some recent analytical work on extending and
generalizing such ideas to more complex systems of equations \cite{ait2008closed, calderon2007fitting, contou2011model, dong2007simplification, givon2004extracting, pavliotis2007parameter,  sotiropoulos2009model}.
%
However, in many instances, closed form, analytical models are not given explicitly,
but can only be inferred from simulation and/or experimental data.
%
We therefore turn to data-driven techniques to analyze such systems and uncover the relevant dynamical modes.
%
In particular, we will use a manifold-learning based approach, as such methods can accommodate nonlinear structures in high-dimensional data.

@@

We present detailed analysis for our method, and provide conditions under which it will successfully recover the slow variables.
%
Furthermore, based on this analysis, we present data-driven protocols to tune the parameters of the method appropriately.
%
Our presentation and discussion addresses two-time-scale stochastic systems; however, we claim that our framework and analysis readily extends to systems with multiple time scale separations.
%

@@

\begin{figure}[t]
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{data_init_nonlinear}
\caption{}
\end{subfigure}
%
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{data_init}
\caption{}
\end{subfigure}
%
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{data_rescaled}
\caption{}
\end{subfigure}
\caption{}
\label{fig:data_rescaling}
\end{figure}

In this work, we assume that we observe data $\mathbf{y}(t_1), \dots, \mathbf{y}(t_N)$, where $t_i = t_0 + i \Delta t$, collected from a single trajectory of a stochastic process $\mathbf{y}(t)$. 
%
We further assume that there exists a nonlinear coordinate transform $\mathbf{f}$ such that $\mathbf{y}(t) = \mathbf{f} \left( \mathbf{x} (t) \right)$, where $\mathbf{x}$ comes from the following autonomous two-time-scale system of stochastic differential equations (SDEs),
\begin{equation} \label{eq:general_SDE}
\begin{aligned}
dx_i(t) &= a_i(\mathbf{x}(t)) dt + dW_i(t), & \: 1 \le i \le m \\
dx_i(t) &= \frac{a_i(\mathbf{x}(t))}{\epsilon} dt + \frac{1}{\sqrt{\epsilon}} dW_i(t) , & \: m+1 \le i \le n
\end{aligned}
\end{equation}
where $W_i(t)$ are independent standard Brownian motions, $\mathbf{x}(t)  = \begin{bmatrix} x_1(t) & \cdots & x_n(t) \end{bmatrix}^T \in \mathbb{R}^n$, and $\epsilon \ll 1$.
%
Here, we restrict ourselves to the case where $\mathbf{x}(t)$ comes from an SDE with additive noise; the case of multiplicative noise is the subject of future research.
%
The parameter $\epsilon$ induces a separation of time scales, such that $\mathcal{O}(1)$ changes in $x_1, \dots, x_m$ occur over times scales $\mathcal{O}(1)$, and $\mathcal{O}(1)$ changes in $x_{m+1}, \dots, x_n$ occur over time scales $\mathcal{O}(\epsilon)$.

Our goal is extract a parameterization of the observed data $\mathbf{y}(t)$ which is consistent with the underlying {\em slow} variables $x_1, \dots, x_m$. 
%
We opt to use a data-driven technique (diffusion maps) which does not directly utilize any information about the underlying dynamics.
%
The main component in the diffusion maps algorithm is defining the relevant pairwise distance metric.
%
Therefore, to analyze dynamic data using diffusion maps, we first need to transform our data to (1) factor out the effects of the (perhaps nonlinear) function $\mathbf{f}$, and (2) rescale the data so that the length scale fluctuations are commensurate with the time scale fluctuations (see Figure~\ref{fig:data_rescaling}). 
%
In this work, we propose a distance metric which incorporates these two transformations in a single step.

Such a rescaling transformation of the SDEs in \eqref{eq:general_SDE} is guaranteed to exist for Khasminskii systems; however, we suspect that our method will be applicable for a broader class of multiscale SDEs. 
%
We are interested in the more general case where we observe $\mathbf{y}(t) = \mathbf{f}( \mathbf{x}(t))$. 
%
In order to factor out the effects of $\mathbf{f}$, we assume that $\mathbf{g} = \mathbf{f} ^{-1}$ is well-defined on the image of $\mathbf{f}$, and both $\mathbf{f}$ and $\mathbf{g}$ be continuously differentiable to fourth order.
%
For the examples presented here, we have $\mathbf{f}: \mathbb{R}^d \mapsto \mathbb{R}^d$; however, we note that in general, the function $\mathbf{f}: \mathbb{R}^n \mapsto \mathbb{R}^d$, $n \le d$, can be a deterministic (possibly nonlinear) function whose image is an $n$-dimensional manifold $\mathcal{M}$ in $\mathbb{R}^d$.
%
We further note that the data must be sampled such that $N \Delta t \gg \epsilon$; this condition ensures that fast variables have equilbrated, and the slow variables have evolved appreciably.

%In the simple case of a linear drift function, i.e., when $a_i(\mathbf{x}(t)) = \mu _i x_i$ with $\mu_i < 0$, the probability density function of $x_i$ approaches a Gaussian with (finite) variance $\mu_i$.
%
%The time constant of the approach of the variance to equilibrium is $-1/\mu_i$ for $i=1,\ldots,m$ and $-\epsilon/\mu_i$ for $i=m+1,\ldots,n$ \cite{lelievre2013optimal}.
%
%Thus, the last $n-m$ variables of \eqref{eq:general_SDE} rapidly approach a local equilibrium measure and exhibit fast dynamics, while the first $m$ variables exhibit slow dynamics.
%
%Because of the separation of time scales in the diffusion terms, a short burst of simulation (with $\delta t = o \left( \epsilon \right)$) will yield a cloud of points which is broadly distributed ($\mathcal{O} \left( 1/\sqrt{\epsilon} \right)$) in the last $n-m$ fast directions but narrowly distributed ($\mathcal{O} \left( 1 \right)$) in the first $m$ slow ones.
%
%With the appropriate conditions on $a_i(\mathbf{x})$, the same can be said for more general drift functions, where $\mu_i$ are the eigenvalues of the Jacobian of $\mathbf{a}(\mathbf{x}) = \begin{bmatrix} a_1(\mathbf{x}) & \cdots & a_n(\mathbf{x}) \end{bmatrix}^T$ \cite{villani2009hypocoercivity}.
%
%Therefore, \eqref{eq:general_SDE} defines an $n$-dimensional stochastic system with $m$ slow variables and $n-m$ fast variables, and $\epsilon$ defines the time scale separation.
%
%The ratio of the powers of $\epsilon$ in the drift and diffusion terms in \eqref{eq:general_SDE} is essential, as we require the square of the diffusivity to be of the same order as the drift as $\epsilon \rightarrow 0$ \cite{berglund2003geometric}.
%
%If the diffusivity is larger, then, as $\epsilon \rightarrow 0$, the equilibrium measure will be unbounded.
%
%Conversely, if the diffusivity is smaller, the equilibrium measure will go to $0$ as $\epsilon \rightarrow 0$.

%Assuming the ensemble average of $a_i(\mathbf{x})$, conditioned on the slow variables $x_1, \dots, x_m$, converges to a distribution which is {\em only} a function of the slow variables, then by the averaging principle \cite{freidlin2012random}, we can write a reduced SDE in {\em only} the slow variables $x_1, \dots, x_m$.
%
%We assume that our data is collected over a time scale much larger than $\epsilon$, such that the $n-m$ fast variables have saturated and we have observed appreciable evolution in the slow variables. 
%
%The aim of our work is to show how we can detect such slow variables {\em automatically} from data, in order to help inform modeling efforts and aid in the writing of such reduced stochastic models.
%
%In general, we are not given the variables $\mathbf{x}(t)$ from the original SDE system, but instead, we are given some {\em observations} in the form $\mathbf{y}(t) = \mathbf{f} (\mathbf{x}(t))$.
%
%We assume that $\mathbf{f}: \mathbb{R}^n \mapsto \mathbb{R}^d$, $n \le d$, is a deterministic (possibly nonlinear) function whose image is an $n$-dimensional manifold $\mathcal{M}$ in $\mathbb{R}^d$.
%
% TODO: check whether the inverse condition is not too strong (and we only need it to apply to the domain of the samples)
%
%For our analysis, we require $\mathbf{g} = \mathbf{f} ^{-1}$ to be well-defined on $\mathcal{M}$, and both $\mathbf{f}$ and $\mathbf{g}$ to be continuously differentiable to fourth order.
%
%Given data $\mathbf{y}(t_1),\ldots,\mathbf{y}(t_N)$ on $\mathcal{M}$ we would like to recover a parametrization of the data that is one-to-one with the slow variables $x_1, \dots, x_m$.

@@

If two points $\mathbf{x}(t_1)$ and $\mathbf{x}(t_2)$ are drawn from an $n$-dimensional
Gaussian distribution with covariance $\mathbf{C}_x$, the Mahalanobis distance between the points is defined as \cite{mahalanobis1936generalized}
\begin{equation}
	\| \mathbf{x}(t_1) - \mathbf{x}(t_2) \| _M = \sqrt{ (\mathbf{x}(t_1) - \mathbf{x}(t_2))^T \mathbf{C}_x^{-1} (\mathbf{x}(t_1) - \mathbf{x}(t_2) )  }.
\end{equation}
In particular, 
%%%YGK+CDS if $\mathbf{x}(t_1)$ and $\mathbf{x}(t_2)$ are samples 
for \eqref{eq:general_SDE}, whose covariance does not depend on $\mathbf{x}$,  $\mathbf{C}_x^{-1} = \mathrm{diag}(e_1, \ldots, e_n)$ is a constant matrix where
\begin{equation} \label{eq:e_def}
\begin{aligned}
e_i =& 1, \: & 1 \le i \le m \\
e_i =& \epsilon, \: & m+1 \le i \le n,
\end{aligned}
\end{equation}
and a short burst of simulation (with $\delta t = o \left( \epsilon \right)$) yields a cloud of points which is broadly distributed ($\mathcal{O} \left( 1/\sqrt{\epsilon} \right)$) in the last $n-m$ fast directions but narrowly distributed ($\mathcal{O} \left( 1 \right)$) in the first $m$ slow ones.
%
The Mahalanobis distance between samples is
\begin{equation} \label{eq:rescale_x_dist}
\| \mathbf{x}(t_2) - \mathbf{x}(t_1) \|^2_M = \sum_{i=1}^n e_i \left( x_i(t_2) - x_i(t_1) \right)^2.
\end{equation}
Note that the metric in \eqref{eq:rescale_x_dist} is insensitive to variations in the fast variables.
%
The metric \eqref{eq:rescale_x_dist} can be rewritten as
\begin{equation} \label{eq:norm_z}
\| \mathbf{x}(t_2) - \mathbf{x}(t_1) \|^2_M = \| \mathbf{z}(t_2) - \mathbf{z}(t_1) \|^2_2
\end{equation}
where
\begin{equation} \label{eq:general_rescale}
z_i(t) = \sqrt{e_i} x_i(t).
\end{equation}
$\mathbf{z}(t)$ is a stochastic process of the same dimension as $\mathbf{x}(t)$, rescaled so that each variable has unit diffusivity.
%
%This rescaling transforms our problem from one of detecting the slow variables within dynamic data to one of traditional data mining.
%
The Mahalanobis distance incorporates information about the dynamics and relevant time scales, so that using traditional data mining techniques with this metric will allow us to detect the slow variables in our data \cite{singer2009detecting}.
%
%It is important to note that, in practice, we {\em never construct} $\mathbf{z}(t)$ {\em explicitly}.
%
It was shown in \cite{singer2008non} that, assuming $\mathbf{f}$ is bilipschitz, the Mahalanobis distance can be extended to approximate (to fourth order) the Euclidean distance between the rescaled samples $\mathbf{z}(t)$ from accessible $\mathbf{y}(t) = \mathbf{f} (\mathbf{x}(t))$,
%
\begin{equation} \label{eq:mahalanobis}
%\| \mathbf{y}(t_2) - \mathbf{y}(t_1) \|^2_M = \| \mathbf{z}(t_2) - \mathbf{z}(t_1) \|^2_2 + \mathcal{O}(\| \mathbf{y}(t_2) - \mathbf{y}(t_1) \|^4_2).
\| \mathbf{y}(t_2) - \mathbf{y}(t_1) \|^2_M = \| \mathbf{z}(t_2) - \mathbf{z}(t_1) \|^2_2 + \mathcal{O}(\| \mathbf{y}(t_2) - \mathbf{y}(t_1) \|^4_2).
\end{equation}
%
Clearly, this approximation is accurate when $\| \mathbf{y}(t_2) - \mathbf{y}(t_1) \|_2^4 \ll \| \mathbf{z}(t_2) - \mathbf{z}(t_1) \|^2_2$; in Section~\ref{subsec:obsfn}, we provide a more detailed analysis of the errors incurred in this approximation.
%
Because we integrate these distances into a manifold learning algorithm which only considers local distances, we recover a parametrization of the data which is consistent with the underlying system variables $\mathbf{x}(t)$, even when the data are obscured by a function $\mathbf{f}$.
%
Section~\ref{subsec:cov_est} shows how we approximate this distance directly from data $\mathbf{y}(t)$.
%

@@

Following \cite{singer2008non}, we show via Taylor expansion that the Mahalanobis distance between the observations \eqref{eq:mahalanobis_distance} approximates the Euclidean distance {\em in the rescaled variables} for general nonlinear observation functions $\mathbf{f}$ (provided $\mathbf{f}$ is bilipschitz and both $\mathbf{f}$ and $\mathbf{f}^{-1}$ are differentiable to fourth order).
%Our approximation of the pairwise distances relies on the Taylor expansion of the measurement function $\mathbf{f}$.
%
\eqref{eq:mahalanobis_distance} cannot be evaluated directly since we do not have access to the covariance matrices, so we will instead estimate the covariances directly from data.
%
We estimate the covariance $\mathbf{C}(\mathbf{y}(t_0))$ empirically from a set of values $\mathbf{y}(t_1), \dots, \mathbf{y}(t_q)$ drawn from the local distribution at $\mathbf{y}(t_0)$, by running $q$ simulations for a short time $\delta t = o (\epsilon) $, each starting from $\mathbf{y}(t_0)$; we refer to this as a {\em simulation burst}.
%
Alternatively, we can consider a single time series of length $q \delta t = o (\epsilon) $ starting from $\mathbf{y}(t_0)$, and then estimate the covariance
from the increments $\Delta \mathbf{y}(t_i) = \mathbf{y}(t_i) -\mathbf{y}(t_{i-1})$.
%
Although we will present analysis and results for the first type of estimation, the second case is often more practical in practice.
%
%It can be shown that if $q \delta t$ is sufficiently small, the errors introduced are of sufficiently low order.

Errors in our estimation of the Mahalanobis distance arise from three sources.
%
One source of error, discussed in Section~\ref{subsec:obsfn}, is approximating the function $\mathbf{f}$ locally as a linear function by truncating the Taylor expansion of $\mathbf{f}$ at first order.
%
An additional source of error, discussed in Section~\ref{subsec:cov_est}, arises from disregarding the drift in the stochastic process, and assuming that samples are drawn from a Gaussian distribution.
%
The third source comes from finite sampling effects, which is the subject of future research.
%
In this work, we address and discuss the first two sources of error (the finite sampling effects are the subject of future research).
%
We control the effects of the errors due to truncation of the Taylor expansion by adjusting $\sigma_{kernel}$; the higher-order terms in this expansion will be small for points which are close, such that adjusting $\sigma_{kernel}$ will allow us to only consider distances which are sufficiently accurate in our overall computation scheme.
%
Furthermore, we control the errors incurred by disregarding the drift by adjusting the time scale of our simulation bursts $\delta t$.
%
Figure~\ref{fig:schematic} illustrates some of the issues in choosing the sizes $\delta t$ (or $q \delta t$ if the alternate method is used) and the parameter $\sigma_{kernel}$.
%
We present both analytical results for the error bounds, as well as an empirical methodology to set the parameters $\sigma_{kernel}$ and $\delta t$ for our method to accurately recover the slow variable(s).

\subsection{Error due to the observation function $\mathbf{f}$} \label{subsec:obsfn}

We want to relate the distance in the rescaled space, $\|\mathbf{z}(t_2) - \mathbf{z}(t_1)\|_2$, to the estimated Mahalanobis distance between the observations $\| \mathbf{y}(t_2) - \mathbf{y}(t_1)\|_M$.
%
We define the error incurred by using the Mahalanobis distance to approximate the true distance as
\begin{equation}
E_M(\mathbf{y}(t_1), \mathbf{y}(t_2)) = \|\mathbf{z}(t_2) - \mathbf{z}(t_1)\|_2^2 - \| \mathbf{y}(t_2) - \mathbf{y}(t_1)\|^2_M .
\end{equation}
%
By Taylor expansion of $\mathbf{g}(y) = \mathbf{f}^{-1}(y)$ around $\mathbf{y}(t_1)$ and $\mathbf{y}(t_2)$ and averaging the two expansions, we obtain
%
\begin{equation} \label{eq:mahanaobis_error}
\begin{aligned}
E_M\left( \mathbf{y}(t_1), \mathbf{y}(t_2) \right)
 =&
\begin{aligned}[t]
 \frac{1}{2} \sum_{i=1}^n \sum_{jkl=1}^{d} &
\left( g_{i, (j)} (\mathbf{y}(t_1)) g_{i, (k,l)} (\mathbf{y}(t_1)) -  g_{i, (j)} (\mathbf{y}(t_2)) g_{i, (k,l)} (\mathbf{y}(t_2)) \right) \times \\
& ({y}_j(t_2) - {y}_j(t_1)) ({y}_k(t_2) - {y}_k(t_1))({y}_l(t_2) - {y}_l(t_1))
\end{aligned} \\
+&
\begin{aligned}[t]
\frac{1}{8} \sum_{i=1}^n \sum_{jklm=1}^d  &
\left( g_{i, (j,k)} (\mathbf{y}(t_1)) g_{i, (l,m)} (\mathbf{y}(t_1)) +  g_{i, (j,k)} (\mathbf{y}(t_2)) g_{i, (l,m)} (\mathbf{y}(t_2)) \right) \times
 \\
&({y}_j(t_2) - {y}_j(t_1))  ({y}_k(t_2) - {y}_k(t_1))({y}_l(t_2) - {y}_l(t_1)) ({y}_m(t_2) - {y}_m(t_1))
\end{aligned} \\
+&
\begin{aligned} [t]
\frac{1}{6} \sum_{i=1}^n \sum_{jklm=1}^d &
\left( g_{i, (j)} (\mathbf{y}(t_1)) g_{i, (k,l,m)} (\mathbf{y}(t_1)) +  g_{i, (j)} (\mathbf{y}(t_2)) g_{i, (k,l,m)} (\mathbf{y}(t_2)) \right) \times \\
& ({y}_j(t_2) - {y}_j(t_1))  ({y}_k(t_2) - {y}_k(t_1))({y}_l(t_2) - {y}_l(t_1))({y}_m(t_2) - {y}_m(t_1))
\end{aligned} \\
+& \mathcal{O} \left(\| \mathbf{y}(t_2) - \mathbf{y}(t_1) \|^6_2 \right) ,
\end{aligned}
\end{equation}
%
where
%
\begin{equation}
\begin{aligned}
g_{i,(j)} &= \sqrt{e_i} \frac{\partial g_i}{\partial y_j}
\\
g_{i,(j,k)} &= \sqrt{e_i}  \frac{\partial^2 g_i}{\partial y_j \partial y_k}
\\
g_{i,(j,k,l)} &= \sqrt{e_i}  \frac{\partial^3 g_i}{\partial y_j \partial y_k \partial y_l} .
\end{aligned}
\end{equation}
%
Singer and Coifman \cite{singer2008non} showed that the error incurred by using the Mahalanobis distance to approximate the $L_2$-distance between points $\mathbf{z}(t)$ is $\mathcal{O} (\|\mathbf{y}_1 - \mathbf{y}_2 \|_2^4 )$ (see the Supplementary Materials for details).
%
We now see from \eqref{eq:mahanaobis_error} that the error is an explicit function of the second- and higher-order derivatives of $\mathbf{g} = \mathbf{f}^{-1}$ and the distance between samples $\| \mathbf{y}(t_2) - \mathbf{y}(t_1) \|_2$.
%
We would like to note that this error does not depend on the dynamics of the underlying stochastic process (as we assume the covariances at each point on the manifold are known), but is only a function of the measurement function $\mathbf{f}$ and the distance between samples $\| \mathbf{y}(t_2) - \mathbf{y}(t_1) \|_2$.
%
The parameter $\sigma_{kernel}$ in the diffusion maps calculation determines how much $E_M$ contributes to the overall analysis.
%
From \eqref{eq:dmaps_kernel}, distances which are much greater than $\sigma_{kernel}$ are negligible in the diffusion maps computation because of the exponential kernel.
%
Therefore, we want to choose $\sigma_{kernel}^2$ on the order of $\|\mathbf{y}(t_2) - \mathbf{y}(t_1)\|^2_M$ in a regime where $| E_M(\mathbf{y}(t_1), \mathbf{y}(t_2))|  \ll \|\mathbf{y}(t_2) - \mathbf{y}(t_1)\|^2_M$.
%
This is illustrated in Figure~\ref{fig:schematic}, where we want to choose $\sigma_{kernel}$ small enough so that the curvature and other nonlinear effects (captured in the error term $E_M$) are negligible.
%
This will ensure that the errors in the Mahalanobis distance approximation do not greatly effect our overall analysis.
